---
title: "719 HW 9"
author: "Will Powell"
date: '2023-11-16'
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. (10 points) Suppose that y|θ∼Gamma(1, θ) (i.e., exponential distribution with mean θ) and
that θ∼InverseGamma(α, β). Find the posterior distribution of θ. Is the inverse gamma a
conjugate prior?

$$p(\theta | y) \propto p(y | \theta) p(\theta)$$
$$\propto \theta e^{-y / \theta} \frac{\beta^{\alpha}}{\Gamma(\alpha)}\theta^{-\alpha-1}exp(-\beta/\theta)$$

$$\propto \frac{\beta^{\alpha}}{\Gamma(\alpha)}\theta^{-\alpha}exp(-(y +\beta)/\theta)$$

Thus $\theta | y \sim InverseGamma(\alpha + 1, \beta + y)$


2. (10 points) Suppose that we have a Normally distributed likelihood $y|\mu_l \sim N(\mu_l, \sigma_l^2)$ and have a prior $y|\mu_l \sim N(\mu_0, \sigma_0^2)$ We observe $y = 0.580$ and set $\mu0 = 0, \sigma_0^2 = 0.1907$ , and $\sigma_l^2 = 0.2266$.
Apply the Metropolis algorithm to generate samples from the posterior distribution of µl.
Use $Q \sim N(0, 1)$. Show a trace plot of the samples. Calculate posterior mean and variance of $\mu_l$ using the samples. Provide your R code.
(Hint: Modify the f.metropolis.x function from lecture)

```{r}
f.metropolis.x <- function(niter, x0) {
  tmp.x <- rep(NA, niter)
  tmp.x[1] <- x0 # choose a starting point
  
  for(ii in 2:niter) {
    tmp.x[ii] <- tmp.x[ii-1] + rnorm(1) # propose new value
    tmp.dist.x <- (1 / sqrt(.1907) ) * exp( (-1 / (2*.1907) )*(tmp.x[ii-1]-0)^2) 
    tmp.dist.x.proposed <- (1 / sqrt(.2266) ) * exp( (-1 / (2*.2266) )*(tmp.x[ii]-0)^2) 
    tmp.alpha <- min(tmp.dist.x.proposed/tmp.dist.x, 1)
    tmp.U <- runif(1)
    tmp.x[ii] <- ifelse( tmp.U < tmp.alpha, tmp.x[ii], tmp.x[ii-1] )
  }
  return(tmp.x)
}
set.seed(10)
x.sam <- f.metropolis.x(niter=50000, x0=.580)
plot(x.sam, type="l")
```

```{r}
post.mean <- mean(x.sam[10001:50000])
post.var <- var(x.sam[10001:50000])
c(post.mean, post.var)
```


Above we can see the calculated posterior mean and variance of $\mu_l$.

